---
title: "The qualitative impact of the storm weather disasters on the population health and economic consequences in the US "
date: "May 3, 2016"
author: "Anton Biryukov, U of Calgary"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary of the analysis (synopsis)

In this brief report I have quantitatively analyzed the property and crop damage caused by the storm weather events, as well as the impact on the population health in the United States. The analyzed data covers the period from 1950 -- 2011, and includes records on major weather events, classified by the event type, with information on the property and crop damage, as well as fatalities/injuries delivered. The preliminary analysis of the documented event types has shown that ~900 types in the raw data can be categorized in 7 following common types. Subsequently, the data was aggregated according to the weather event type. For each event type, I computed the total economy damage ( property + crop), as well as count of injuries and fatalities, and segregated that over the states. The results demonstrate the fact that tornadoes tend to produce the largest impact on the population health as measured by the count of fatalities and injuries across all the states. On the other hand, the greatest economic consequences are somewhat equally caused by heavy-wind related disasters, tornadoes and hails. The results also show that the consequences of floods are considerable, whereas the economic footprint of winter seasonal events is relatively small.

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The dataset offered for the analysis in this study is relatively dirty, and needs some preprocessing prior to being able to assist in answering the research questions. Specifically, the research questions are:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

These two questions imply the event types to be clearly outlined in the dataset, although that is not true. Therefore, below I explain the processing done to categorize the events into several types.

## Data Processing
### Libraries Preloading
I start off by loading the necessary libraries for the analysis, as well as downloading the data itself
```{r}
# Load the necessary libraries:
library("dplyr")
library("lubridate")
library("ggplot2")
library("R.utils")
# This function downloads data if necessary:
source("./downloadData.R")
downloadData()
```
### Reading the data
Now, when data is downloaded, I check whether it is already in the memory. if not - I read the csv with the data, otherwise I proceed further towards analysis. As data is really ``heavy'', it totally makes sense to have this check in the script.
```{r, cache=TRUE}
if (!exists("stormData") )
{
    # Unzip the data
    if (!file.exists("./Data/Storm_dataset.csv")){
        message("unzipping the data")
        bunzip2("./Data/StormData.csv.bz2", "./Data/Storm_dataset.csv", remove = FALSE, skip = TRUE)
        message("Done unzipping!")
        
    }
    message("Forming a subset of data")
    if (work_on_subset){
        stormData <- tbl_df(read.csv("./Data/Storm_dataset.csv",nrows = n_row))
    }
    else{
        stormData <- tbl_df(read.csv("./Data/Storm_dataset.csv"))
    }
}
```
### Correcting the factors for property dmg
According to the documentation, the property and the crop damage field are followed by the exponential multiplier (K,M,B), indicating thousands , millions and billions of dollars, respectively. Therefore, this function kmb_to_exp will take care of the conversion from k,m,b to 1E3, 1E6, 1E9.

```{r}
# Establish the mapping from K and M to 10^3 and 10^6
exp_val <- c(1E3,1E6,1E9)
names(exp_val) <- c("K","M","B")  # All numbers are then in $$$
kmb_to_exp <- function(x,exp_val){
    
    if (toupper(x) %in% names(exp_val)) y <- as.numeric(exp_val[toupper(x)])
    else y <- NA
}    

```

Now I add the columns that would transform letters to the numeric values, as well as filter out NAs:

```{r, cache = TRUE}
storm_Data_PROP <- stormData %>%
    mutate(PROPDMGEXP_ADJ = sapply(PROPDMGEXP,kmb_to_exp,exp_val)) %>%
    mutate(CROPDMGEXP_ADJ = sapply(CROPDMGEXP,kmb_to_exp,exp_val) ) %>%
    filter(!is.na(PROPDMG) & !is.na(PROPDMGEXP_ADJ))

```
### Categorizing the EVTYPE column
The reason to categorize the column is that it has A LOT of unique values:
```{r}
length(unique(stormData$EVTYPE))
```
Here I show top-20 most frequent types and their share in the total:
```{r}
typesStats <- sort(table(stormData$EVTYPE),decreasing = TRUE)
head(typesStats,20)
percentageTop <- sum(typesStats[1:20])/sum(typesStats)
print(percentageTop)
```
Seems like this is >95% of data, therefore I can ground my segregation based on the top 20 types. It seems that I could potentially divide the 20 into 6-7, e.g. with the following keywords. These keywords would then serve as a event type identifier and a regular expression. Therefore, I go over the records, and try to find one of the 7 words in its EVTYPE string, and then assign it the type according to the keyword it includes:
```{r, cache = TRUE}
keywords =c("HAIL","WIND","TORNADO|FUNNEL","FLOOD|FLD","LIGHTNING","SNOW|WINTER|ICE","RAIN|WATER")
strings <- storm_Data_PROP$EVTYPE
table_greps <- sapply(keywords, grepl, strings, ignore.case=TRUE)
message("Looking for the keywords in the EVTYPE")
evtypes_df <- arrange(tbl_df(as.data.frame(which(table_greps==TRUE,arr.ind = TRUE ))),row) %>% mutate(EVtype = keywords[col])
Evtypes <- array(rep(NA,length(strings)))
Evtypes[as.array(evtypes_df$row)] = evtypes_df$EVtype # Assign subsetted event type according to the row number
```
Now I am ready to analyze the relative impact of the aforementioned types:

```{r,cache = TRUE}
storm_Data_Table <- storm_Data_PROP %>%
    mutate(Total_Dmg_Corrected = sum(c(PROPDMG*PROPDMGEXP_ADJ,CROPDMG*CROPDMGEXP_ADJ),na.rm=TRUE)) %>% # Apply the exp correction and sum the impact
    mutate(eventType = Evtypes) %>% # Aggregate the types of the events into 7 subtypes
    filter(!is.na(eventType)) %>%
    filter(STATE %in% state.abb) %>% # Make sure the abbreviations have the states as listed in state.abb
    group_by(STATE,eventType) %>% # Group by the event type and the state
    summarize(total_dmg = sum(Total_Dmg_Corrected)) # Sum the damage per event type
```







## Results

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

